---
title: "NiftiArray: Fast Random Access of NIfTI Objects"
output: 
    rmarkdown::html_document:
      base_format: rmarkdown::html_vignette
      keep_md: true
vignette: >
  %\VignetteIndexEntry{Using NiftiArray}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  echo = TRUE,
  fig.width=7, 
  fig.height=5,
  fig.align = 'center'
)
```

This page is still under construction! Check back later for updates!

# Overview

`R` is not well suited for big data sets. NIfTI images, depending on the image dimension and voxel size, can be quite large when loaded and memory is a concern. As sample size or number of scans increases it becomes difficult to perform simple operations voxel-wise across subjects. For example, calculating the mean image across 800 subjects is computationally intense since not all 800 subjects can be loaded into memory at once. Therefore, there is a need for alternative approaches. 

The `NiftiArray` package allows for fast [random access](https://en.wikipedia.org/wiki/Random_access) 
of imaging data in NIfTI format and supports [`DelayedArray` operations](https://bioconductor.org/packages/release/bioc/html/DelayedArray.html). The package establishes the `NiftiArray` class, a convenient and memory-efficient array-like container for on-disk representation of NIfTI image(s). The `NiftiArray` class is an extension of the [`HDF5Array` class](https://bioconductor.org/packages/release/bioc/html/HDF5Array.html) and converts NIfTI objects on disk to HDF5 files which allow for block processing and memory-efficient representations in R.

`NiftiArray` is compatible with the `DelayedArray` and `DelayedMatrixStats` packages.

`DelayedArray` is an R package currently [hosted on Bioconductor](https://bioconductor.org/packages/release/bioc/html/DelayedArray.html). `DelayedArray` allows common array operations on an object without loading it into memory. In order to reduce memory usage and optimize performance, operations on the object are either delayed or executed using a block processing mechanism. `DelayedArray` works with the `NiftiArray` class to delay computation and allow fast and efficient data access.

`DelayedMatrixStats` is an R package currently [hosted on Bioconductor](https://bioconductor.org/packages/release/bioc/html/DelayedMatrixStats.html). `DelayedMatrixStats` contains functions for statistical calculations (i.e. row or column median calculation) using `DelayedArray` efficient block processing on large matrices while keeping local memory usage low. `DelayedMatrixStats` builds on both the `DelayedArray` and `matrixStats` packages to allow for  high-performing functions operating on rows and columns of `DelayedMatrix` objects. The functions are optimized by data type and for subsetted calculations such that both memory usage and processing time is minimized.

<!--
TODO: Add this section somewhere
The Hierarchical Data Format version 5 (HDF5), is a file format that supports large, complex, heterogeneous data. This file format allows for fast random access of data on-disk. That is, HDF5 files can be directly manipulated without loading the object into memory and overcome a number of the issues related to working with big data in R. 

The `rhdf5` R package allows for direct manipulation of the HDF5 dataset on disk. This package can be utilized to convert, read, write, and manipulate the HDF5 data stored on disk. Other packages like `HDF5Array` provide memory-efficient on-disk access to the HDF5 dataset for standard R matrix/array manipulation. Simply put, as a user you feel as though you are working with a locally loaded large array (or matrix) and interact the same way you would if the object was loaded. For example, if you want to extract the value at array index 1,29,300 you can still simply call `data[1,29,.300]`. This large array `data` is never actually loaded into memory but rather is being accessed on disk in real time so memory is kept low and efficient.

Current methods were not designed for use explicitly with 3D array or images but support. So we extended and built something to seamlessly work with images....

This package allows users to save NIfTI objects as HDF5 files while conserving the array structure and NIfTI header. Once the data are stored as an HDF5 they can be represented as a `NiftiArray` object. 

TODO:  Add a GRAPHIC OF NIFTI TO NIFTIARRAY TO MATRIX TO CBINDED MATRIX TO DELAYED TOOLS. PUT IDEA OF MEMORY WITH IT.
-->

# Installation

You can install the development version of `NiftiArray` from [GitHub](https://github.com/) using the following:

```{r}
# install.packages('remotes')
remotes::install_github("muschellij2/NiftiArray")
```

We are working to get a stable version on [Neuroconductor](www.neuroconductor.org).

# Tutorial

## Packages

The packages you will need to load for use with this tutorial are below:

```{r, warning = FALSE, message = FALSE}
library(DelayedArray)
library(DelayedMatrixStats)
library(dplyr)
library(forcats)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(microbenchmark)
library(NiftiArray)
library(neurobase)
library(pbapply)
library(profmem)
library(remotes)
library(RNifti)
library(stringr)
library(tibble)
library(tidyr)
```

## Data

<!-- TODO: Melissa put a better description here. Actually describe the data maybe?-->

This tutorial will use data found [here](https://github.com/muschellij2/open_ms_data/tree/master/cross_sectional/MNI). A description of the data is available [here](https://johnmuschelli.com/open_ms_data/#process-registered-to-mni). In this tutorial we will use the FLAIR images for subjects 1-5. We will download the data to a [temporary folder](https://en.wikipedia.org/wiki/Temporary_folder). If you prefer to load these to a specific directory feel free to change the `nii_destination` in `fileinfo` to the file path where you'd like to save each image.

```{r}
# Information about URL to download and where to save the image locally in destiation
fileinfo = tibble::tibble(url = c('https://raw.githubusercontent.com/muschellij2/open_ms_data/master/cross_sectional/MNI/patient01/T1_N4_noneck_reduced_winsor_regtoFLAIR_brain_N4_regtoMNI.nii.gz',
                                  'https://raw.githubusercontent.com/muschellij2/open_ms_data/master/cross_sectional/MNI/patient02/T1_N4_noneck_reduced_winsor_regtoFLAIR_brain_N4_regtoMNI.nii.gz',
                                  'https://raw.githubusercontent.com/muschellij2/open_ms_data/master/cross_sectional/MNI/patient03/T1_N4_noneck_reduced_winsor_regtoFLAIR_brain_N4_regtoMNI.nii.gz',
                                  'https://raw.githubusercontent.com/muschellij2/open_ms_data/master/cross_sectional/MNI/patient04/T1_N4_noneck_reduced_winsor_regtoFLAIR_brain_N4_regtoMNI.nii.gz',
                                  'https://raw.githubusercontent.com/muschellij2/open_ms_data/master/cross_sectional/MNI/patient05/T1_N4_noneck_reduced_winsor_regtoFLAIR_brain_N4_regtoMNI.nii.gz'),
          nii_destination = c(tempfile(pattern = "flairpatient01", tmpdir = tempdir(), 
                                       fileext = ".nii.gz"),
                              tempfile(pattern = "flairpatient02", tmpdir = tempdir(), 
                                       fileext = ".nii.gz"),
                              tempfile(pattern = "flairpatient03", tmpdir = tempdir(), 
                                       fileext = ".nii.gz"),
                              tempfile(pattern = "flairpatient04", tmpdir = tempdir(), 
                                       fileext = ".nii.gz"),
                              tempfile(pattern = "flairpatient05", tmpdir = tempdir(), 
                                       fileext = ".nii.gz")),
          hdf5_destination = c(tempfile(pattern = "flairpatient01", tmpdir = tempdir(), 
                                        fileext = ".h5"),
                               tempfile(pattern = "flairpatient02", tmpdir = tempdir(), 
                                        fileext = ".h5"),
                               tempfile(pattern = "flairpatient03", tmpdir = tempdir(), 
                                        fileext = ".h5"),
                               tempfile(pattern = "flairpatient04", tmpdir = tempdir(), 
                                        fileext = ".h5"),
                               tempfile(pattern = "flairpatient05", tmpdir = tempdir(), 
                                        fileext = ".h5")))

# Download all the files to the nii_destination
mapply(function(x,y) download.file(url = x, destfile = y), fileinfo$url,fileinfo$nii_destination)

# Notice the files were saved to a temporary directory on your machine
# Change tempdir() to the directory where you saved the files if you adapted nii_destination
list.files(tempdir(), full.names = TRUE)
```

```{r, echo = FALSE}
if(stringr::str_detect(getwd(), '/Users/alval') == TRUE){
    # Information about URL to download and where to save the image locally in destiation
fileinfo = tibble::tibble(url = c('https://raw.githubusercontent.com/muschellij2/open_ms_data/master/cross_sectional/MNI/patient01/T1_N4_noneck_reduced_winsor_regtoFLAIR_brain_N4_regtoMNI.nii.gz',
                                  'https://raw.githubusercontent.com/muschellij2/open_ms_data/master/cross_sectional/MNI/patient02/T1_N4_noneck_reduced_winsor_regtoFLAIR_brain_N4_regtoMNI.nii.gz',
                                  'https://raw.githubusercontent.com/muschellij2/open_ms_data/master/cross_sectional/MNI/patient03/T1_N4_noneck_reduced_winsor_regtoFLAIR_brain_N4_regtoMNI.nii.gz',
                                  'https://raw.githubusercontent.com/muschellij2/open_ms_data/master/cross_sectional/MNI/patient04/T1_N4_noneck_reduced_winsor_regtoFLAIR_brain_N4_regtoMNI.nii.gz',
                                  'https://raw.githubusercontent.com/muschellij2/open_ms_data/master/cross_sectional/MNI/patient05/T1_N4_noneck_reduced_winsor_regtoFLAIR_brain_N4_regtoMNI.nii.gz'),
          nii_destination = c('/Users/alval/Box/rpackages/NiftiArray_examples/data/nii/flairpatient01.nii.gz',
                              '/Users/alval/Box/rpackages/NiftiArray_examples/data/nii/flairpatient02.nii.gz',
                              '/Users/alval/Box/rpackages/NiftiArray_examples/data/nii/flairpatient03.nii.gz',
                              '/Users/alval/Box/rpackages/NiftiArray_examples/data/nii/flairpatient04.nii.gz',
                              '/Users/alval/Box/rpackages/NiftiArray_examples/data/nii/flairpatient05.nii.gz'),
          hdf5_destination = c('/Users/alval/Box/rpackages/NiftiArray_examples/data/hdf5/flairpatient01.h5',
                              '/Users/alval/Box/rpackages/NiftiArray_examples/data/hdf5/flairpatient02.h5',
                              '/Users/alval/Box/rpackages/NiftiArray_examples/data/hdf5/flairpatient03.h5',
                              '/Users/alval/Box/rpackages/NiftiArray_examples/data/hdf5/flairpatient04.h5',
                              '/Users/alval/Box/rpackages/NiftiArray_examples/data/hdf5/flairpatient05.h5'))
}
temp = neurobase::readnii(fileinfo$nii_destination[1])
```




These data are $182 \times 218 \times 182$ with pixel dimension $1 mm \times 1mm \times 1 mm$. On disk they take up approximately 5MB as nii.gz compressed objects. When loaded into R, each image consumes approximately 55 MB worth of memory.  

<!-- TODO: Melissa fill these in.. something generally average... and make the last sentence of the previous paragraph.
On an average computer with xx memory only approximate xx images could be loaded into R simultaneously before crashing.
-->

Obviously, imaging data can be smaller or larger in memory depending on dimension and pixel size but naturally all imaging data will run into memory restrictions. `NiftiArray` attempts to overcome these problems by converting NIfTI objects to HDF5 files, an efficient random access data type on disk. `NiftiArray` conserves the array structure and NIfTI header. Therefore, interacting with `NiftiArray` objects in R is similar to regular NIfTI objects. The benefit here is the images remain on disk and are not ever fully loaded so memory is conserved. See __Comparison of `NiftiArray` to Traditional Methods__ section for memory comparisons.

## Single NIfTI Image to a `NiftiArray`

In the __Data__ section we downloaded data for this tutorial into a temporary directory and recorded the file paths in which they were saved. These files are .nii.gz file types. Unfortunately, these NIfTI compressed images cannot be fast random accessed. In order to utilize the `NiftiArray` functionality these data must be converted to HDF5 files on disk. This will not affect how you interact with the object in R only how it is stored on disk. Unfortunately, saving this objects out as HDF5 objects will require the same amount of memory on disk so if you keep the NIfTI images on disk as well as the HDF5 files you will double your memory usage on disk. In R though, these objects will use almost no memory. This is where `NiftiArray` shines.

### `NiftiArray::writeNiftiArray`

To convert a NIfTI object to an HDF5 file and eventually a local `NiftiArray` object you can use the `NiftiArray::writeNiftiArray` function. By default this function will store the data in a temporary folder, similar to where we saved the data for this tutorial. You can over-ride this though by specifying the `filepath` option.

__Note__: When calling `NiftiArray::writeNiftiArray` you are converting the NIfTI object on disk to a HDF5 file with a NIfTI-specific grouping or hierarchical format. See [this link](https://support.hdfgroup.org/HDF5/doc/H5.intro.html) for more information on groups within HDF5 files. In the event that users have stored other information in groups inside the HDF5 file output from `NiftiArray::writeNiftiArray` we did not want to support over-writing the file and therefore additional information. Instead, we simply over-write the groups that are associated with a `NiftiArray` object inside the HDF5 file. To over-write these groups, set `overwrite = TRUE` in the `NiftiArray::writeNiftiArray` function.

Let's convert and write the first subject's data as a temporary file to the temporary directory on-disk. This temporary file will have the pattern `NiftiArraypatient01` in the file name. Again, if you'd like to save this object somewhere other than the temporary directory simply change the `filepath` option in `NiftiArray::writeNiftiArray`.

```{r}
# Write the NiftiArray object to disk and load it into R as patient01
patient01 = NiftiArray::writeNiftiArray(fileinfo$nii_destination[1], filepath = fileinfo$hdf5_destination[1], overwrite = TRUE)
# List the files in the temporary directory
# Change tempdir() to the directory where you saved the files if you adapted hdf5_destination
list.files(tempdir())
```

The NIfTI was converted to the the NIfTI-HDF5 file format and saved on disk. It was also loaded into memory is a `NiftiArray` object and returned in R as `patient01`. `patient01` is of class `NiftiArray`. 

```{r}
class(patient01)
```

The NIfTI header is conserved in the `NiftiArray` class in case you ever need to quality control or convert a `NiftiArray` back to a NIfTI image. You can extract the header information from a `NiftiArray` object using the `NiftiArray::nifti_header` function.

```{r}
# Print the NIfTI header associated with patient01
NiftiArray::nifti_header(patient01)
```

In the next section we will re-load `patient01` using the `NiftiArray::NiftiArray` function. Let's remove the `patient01` object from R memory so we can re-load it later.

```{r}
# Remove patient01 object from R memory so it can be re-loaded later
rm(patient01)
```

The `NiftiArray::writeNiftiArray` function takes a single file path and converts the NIfTI image on disk to the HDF5 file on disk while also loading the `NiftiArray` into local memory. By default, the images are saved to a temporary directory. This is useful for quick one time calculations or conversions and ensures that on-disk storage remains minimal since the temporary directories delete the files systematically over time. 

When you want to work with `NiftiArray` objects for lots of subjects frequently though the conversion process may take some valuable compute time and memory. Instead, in these instances if on-disk storage is not an issue we suggest saving the HDF5 `NiftiArray` objects to their own folder and loading `NiftiArray` into R using `NiftiArray::NiftiArray`. For only a few images compute time for both approaches (`NiftiArray::writeNiftiArray` and `NiftiArray::NiftiArray`) are fast and comparable but for larger data sets of reasonable dimension and pixel size it can take a few minutes to convert and save the HDF5 object and a few minutes to load into R locally. See the __Compute Time__ section for more details about differences in compute time between `NiftiArray::writeNiftiArray` and `NiftiArray::NiftiArray`.

### `NiftiArray::NiftiArray`

The `NiftiArray` function can be used to load an on disk NIfTI object into R as a `NiftiArray`. 

```{r}
# Reload patient01 using NiftiArray::NiftiArray from a NIfTI object on disk
patient01 = NiftiArray::NiftiArray(fileinfo$nii_destination[1])
# Remove patient01 object from R memory so it can be re-loaded later
rm(patient01)
```

It can also load the `NiftiArray` object from the HDF5 file converted using `NiftiArray::writeNiftiArray` object on-disk and stored as a HDF5 it can be loaded into R using the `NiftiArray::NiftiArray` function. 

```{r}
# Reload patient01 using NiftiArray::NiftiArray from a HDF5 object on disk
patient01 = NiftiArray::NiftiArray(fileinfo$hdf5_destination[1])
```

Notice this is exactly the same `patient01` object as before. 

### Interacting with `NiftiArray` Objects

Though a `NiftiArray` object you'll interact with R and the `NiftiArray` as you would a normal image. 

```{r}
# Index a single voxel
patient01[90, 89, 101]
# Change the value of the background voxels from 0 to -100
## Note: Normally this should be done using a brain mask but this is just an example for this tutorial
patient01[patient01 == 0] = -100
# Summary stats
min(patient01)
max(patient01)
# Image operations
patient_sum = patient01 + patient01
# Notice it did sum the NiftiArray objects voxel-wise across objects
patient_sum[90, 89, 101]
```

Due to the block processing some functionality will not be available.

```{r}
# Due to the blcok calculation some base functions will not be available
# The table function errors
# table(patient01)
```

```{r}
# Remove patient01 object from R memory so it can be re-loaded later
rm(patient01)
```

## Multiple NIfTI Images to `NiftiArray` Objects

In practice we have a list of subjects we want to convert from NIfTIs to `NiftiArray` objects and load into R. 

### `mapply` `NiftiArray::writeNiftiArray`

<!-- 
TODO: Fix NiftiArrayList to properly call lapply when given an nii to convert and a specific file path to convert it to. Then remove this section and add it to the NiftiArrayList code.
-->

To write out multiple NIfTI images on disk to the required HDF5 files we can simply `mapply` over `NiftiArray::writeNiftiArray`.

```{r}
# Write out the entire filepath of NIfTI objects to HDF5s
all_patients_list = mapply(function(x,y) writeNiftiArray(x = x, filepath = y, overwrite = TRUE), fileinfo$nii_destination, fileinfo$hdf5_destination)
# List the files in the temporary directory -- all subjects should be there as NIfTIs and .h5s
list.files(tempdir())
# Remove all_patients_list since we will load it in a different way later
rm(all_patients_list)
```

### `NiftiArray::NiftiArrayList`

The `NiftiArray::NiftiArrayList` function converts and writes `NiftiArray` objects if the `x` input is of class NIfTI and then loads all the `NiftiArray` objects into R in a `list` as a new class `NiftiArrayList`. That is, every element in the list is a `NiftiArray`. 

```{r}
# Convert, write to temporary disk, and load all_patients_list using NiftiArray::NiftiArrayList
all_patients_list = NiftiArray::NiftiArrayList(fileinfo$nii_destination)
# Show the class is NiftiArrayList
class(all_patients_list)
```

The `NiftiArray::NiftiArrayList` class also simply loads the `NiftiArray` objects as a `list` if the `x` input is a set of file paths to the HDF5 `NiftiArray` files on disk.

```{r}
# Load all_patients_list using NiftiArray::NiftiArrayList
all_patients_list = NiftiArray::NiftiArrayList(fileinfo$hdf5_destination)
# Show the class is NiftiArrayList
class(all_patients_list)
```

At this point, we have all 5 patients loaded into R as a `NiftiArrayList` object. We can now convert the `NiftiArray` object to a `NiftiMatrix` object in order to run voxel-wise calculations.

<!--
### Working with a NiftiArrayList
TODO Melissa if possible put function operations for accessing voxels or info across the elements of the list
-->

## Single `NiftiArray` to a `NiftiMatrix`

The `NiftiArray` object is a 3 dimensional array structure that allows for memory efficient delayed random access of NIfTI objects. The `NiftiMatrix` is the result of concatenating the `NiftiArray`. Similar to `NiftiArray`, `NiftiMatrix` is a new class object. Rather than an array structure we can strung out the image to a vector. In the code below, we convert a `NiftiArray` to a `NiftiMatrix` for one patient. We then verify that the class of this object is in fact a `NiftiMatrix`, has only a single column, index the vector to print some values, and validate that the object size is as memory efficient as the `NiftiMatrix`.

```{r} 
# Reload patient01 using NiftiArray::NiftiArray from a HDF5 object on disk
patient01 = NiftiArray::NiftiArray(fileinfo$hdf5_destination[[1]])
# Convert the NiftiArray to a NiftiMatrix
patient01_niimat = as(patient01, "NiftiMatrix")
# Verify patient01_niimat is of class NiftiMatrix
class(patient01_niimat)
# Notice the NiftiMatrix has only 1 column
dim(patient01_niimat)
# Index the NiftiMatrix like a vector
patient01_niimat[40000:40010]
# The NiftiMatrix object is still memory efficient
object.size(patient01_niimat)
# Obtain the NIfTI header
NiftiArray::nifti_header(patient01_niimat)
```

In this example, we showed the result of converting a single patients `NiftiArray` to a `NiftiMatrix` but it will be more useful to create a `NiftiMatrix` with multiple subjects. 

## `NiftiArrayList` to a Big `NiftiMatrix`

In order to use tools like `DelayedArray` and `DelayedMatrixStats` to calculate voxel-level statistics across multiple subjects we need to create a big `NiftiMatrix`. That is, each row will represent a voxel and each column a new subject. To do this, we can take advantage of the `NiftiArrayList` class.

```{r}
# Show all_patients_list is of NiftiArrayList
class(all_patients_list)
# Convert each element in the list to a NiftiMatrix
all_patients_niimat = pbapply::pblapply(all_patients_list, as, "NiftiMatrix")
# Concatenate the list elements column-wise
all_patients_niimat  = do.call(DelayedArray::acbind, all_patients_niimat)
# Show the matrix
head(all_patients_niimat)
# Notice the dimension is all the voxels in the brain by the 5 patients we are working with
dim(all_patients_niimat)
# Obtain the NIfTI header
NiftiArray::nifti_header(all_patients_niimat)
```


### `DelayedMatrixStats`

The `DelayedMatrixStats` package allows for row or column wise statistical operations using delayed block processing to keep both memory and speed optimized. For more information, the package and documentation is available through [Bioconductor here](https://bioconductor.org/packages/release/bioc/html/DelayedMatrixStats.html).

Below we show a simple example obtaining the voxel level mean and median across subjects.

```{r}
# Calculate the voxel-wise  median across subjects
voxel_medians = DelayedMatrixStats::rowMedians(all_patients_niimat)
# Index the median vector
voxel_medians[1043001]
# Show the class of voxel_medians
class(voxel_medians)
# Calculate the object size
object.size(voxel_medians)
```

Notice the resulting vector `voxel_medians` is not a `DelayedArray` or `NiftiArray` object but rather a normal `vector`. We can convert it to a `NiftiMatrix` so that it returns to a memory efficient object using `as`.

```{r}
# Convert the median vector back to a `NiftiMatrix`
voxel_medians = as(voxel_medians, 'NiftiMatrix')
# Notice the size is back to a memory efficient value
object.size(voxel_medians)
# The header is no longer accurate though
NiftiArray::nifti_header(voxel_medians)
```

Notice the NIfTI header is no longer accurate when we coerce the normal vector to a `NiftiMatrix`.

```{r}
# Calculate the voxel-wise  mean across subjects
voxel_means = DelayedMatrixStats::rowMeans2(all_patients_niimat)
```

`DelayedMatrixStats` is a very powerful package so you should spend some time looking through the functions and documentation to see what is available and useful for you.

<!--
TODO: Add example about making voxel wise scatter plots
### Scatter plots?
-->

### Converting Between `NiftiMatrix`, `NiftiArray`, and `NiftiImage` Classes

So long as the object belongs to the classes available in `NiftiArray` (i.e. `NiftiArray`, `NiftiMatrix`) we can easily convert between the classes and back to a `niftiImage` class exported from `RNifti`. It is easy to convert between these object because the NIfTI header is retained among all the objects.

#### NiftiMatrix to NiftiArray

```{r}
class(patient01_niimat)
niimat2niiarr = as(patient01_niimat, "NiftiArray")
class(niimat2niiarr)
```

#### `NiftiMatrix` to `niftiImage`

```{r}
class(patient01_niimat)
niimat2nii = as(patient01_niimat, "niftiImage")
class(niimat2nii)
```

#### `NiftiArray` to `NiftiMatrix`

```{r}
class(patient01)
niiarr2niimat = as(patient01, "NiftiMatrix")
class(niiarr2niimat)
```

#### `NiftiArray` to `niftiImage`

```{r}
class(patient01)
niiarr2nii = as(patient01, "niftiImage")
class(niiarr2nii)
```

### `DelayedMatrixStats` and Vectors to `NiftiArray`

Converting between the objects returned from `DelayedMatrixStats` functions and `NiftiArray` objects is not as easy because the NIfTI header was lost in the calculations involved in `DelayedMatrixStats` functions. We must extract the header from previous object and then re-initialize the `NiftiArray`.

```{r}
# Median
# Obtain the image header associated with the all_patients_list NiftiArrayList
image_header = NiftiArray::nifti_header(all_patients_list)
# Initialize an array
median_arr = array(voxel_medians, dim = image_header$dim[2:4])
# The object size of median_arr is normal because it is a normal array
object.size(median_arr)
# Convert the array to a NiftiArray and write it out to a temporary file
median_niiarr = NiftiArray::writeNiftiArray(median_arr, header = image_header)
# The object size is back to our efficient NiftiArray
object.size(median_niiarr)

# Mean
# Obtain the image header associated with the all_patients_list NiftiArrayList
image_header = NiftiArray::nifti_header(all_patients_list)
# Initialize an array
mean_arr = array(voxel_means, dim = image_header$dim[2:4])
# The object size of mean_arr is normal because it is a normal array
object.size(mean_arr)
# Convert the array to a NiftiArray and write it out to a temporary file
mean_niiarr = NiftiArray::writeNiftiArray(mean_arr, header = image_header)
# The object size is back to our efficient NiftiArray
object.size(mean_niiarr)
```

Once a `NiftiArray` it is easy to create the `niftiImage` object.

```{r}
median_nii = as(median_niiarr, "niftiImage")
mean_nii = as(mean_niiarr, "niftiImage")
```

```{r}
neurobase::ortho2(median_nii)
neurobase::ortho2(mean_nii)
```

You could write these objects out as NIfTIs using `RNifti::writeNifti`

# Comparison of `NiftiArray` to Traditional Methods

## Memory

### Local Object Size

```{r, echo = TRUE}
# Calculate the a single patient object size using different NIfTI read functions
memory = tibble::tibble(read_type = c('NiftiArray::NiftiArray', 
                                      'NiftiArray::writeNiftiArray', 
                                      'RNifti::readNifti', 
                                      'neurobase::readnii'),
                        byte_size = c(object.size(NiftiArray::NiftiArray(fileinfo$nii_destination[1])),
                                      object.size(NiftiArray::writeNiftiArray(fileinfo$nii_destination[1])),
                                      object.size(RNifti::readNifti(fileinfo$nii_destination[1])),
                                      object.size(neurobase::readnii(fileinfo$nii_destination[1]))),
                       log_byte_size = log(byte_size)) %>% 
  dplyr::mutate(read_type = as.factor(read_type),
                read_type = forcats::fct_reorder(read_type, byte_size, .desc = TRUE))

# Table of memory
knitr::kable(
  memory,
  col.names = c('Read Function', 'Byte Size', 'log(Byte Size)'),
  format = 'html',
  digits = 2, 
  caption = 'Memory mapped in bytes from a single patients image read. 
  The image dimension is 182 by 218 by 182 with pixel dimension 1 mm by 1 mm by 1 mm. 
  On disk the image is 4.9 MB.', 
  booktabs = TRUE
) %>% 
  kableExtra::kable_styling("striped", full_width = FALSE) 
```

```{r}
# Bar graph in bytes
ggplot(memory, aes(x = read_type, y = byte_size, fill = read_type)) +
  geom_bar(stat="identity") +
  geom_text(aes(label=byte_size), vjust=-0.3, size=3.5) +
  ylab('Memory in Bytes') +
  xlab('Read Function') +
  labs(title = 'Memory Comparison for Reading a Single NIfTI Image',
       subtitle = 'Image Dimension: 182 by 218 by 182, Pixel Dimension: 1 mm by 1 mm by 1 mm') +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5), legend.position = 'none')

# Bar graph in log(bytes)
ggplot(memory, aes(x = read_type, y = log_byte_size, fill = read_type)) +
  geom_bar(stat="identity") +
  geom_text(aes(label=round(log_byte_size), digits = 2), vjust=-0.3, size=3.5) +
  ylab('Memory in log(Bytes)') +
  xlab('Read Function') +
  labs(title = 'Memory Comparison for Reading a Single NIfTI Image on the Log Scale',
       subtitle = 'Image Dimension: 182 by 218 by 182, Pixel Dimension: 1 mm by 1 mm by 1 mm') +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5), legend.position = 'none')
```

<!--
# TODO: Melissa put these plots together -- use cowplot or gridExtra to clean up
-->


### Memory Profile

<!-- 
TODO Melissa add a scatter plot and line graph using the memory profile for each -- this is especially useful to show that writeNiftiArray uses more memory in the profile but the object is the same 
-->

```{r}
# Initalize a list to store the memory profile
memory_profile = list()
# Memory profile of RNifti::readNifti()
memory_profile[[1]] = tibble::tibble(read_type = 'RNifti::readNifti', 
                                     memory_profile = profmem::profmem(RNifti::readNifti(fileinfo$nii_destination[1]))$bytes)
memory_profile[[2]] = tibble::tibble(read_type = 'neurobase::readnii', 
                                     memory_profile = profmem::profmem(neurobase::readnii(fileinfo$nii_destination[1]))$bytes)
memory_profile[[3]] = tibble::tibble(read_type = 'NiftiArray::NiftiArray - NIfTI File', 
                                     memory_profile = profmem::profmem(NiftiArray::NiftiArray(fileinfo$nii_destination[1]))$bytes)
memory_profile[[4]] = tibble::tibble(read_type = 'NiftiArray::writeNiftiArray', 
                                     memory_profile = profmem::profmem(NiftiArray::writeNiftiArray(fileinfo$nii_destination[1]))$bytes)
memory_profile[[5]] = tibble::tibble(read_type = 'NiftiArray::NiftiArray - HDF5 File',
                                     memory_profile = profmem::profmem(NiftiArray::NiftiArray(fileinfo$hdf5_destination[1]))$bytes)

memory_profile = dplyr::bind_rows(memory_profile) %>% 
  tidyr::drop_na() %>% 
  dplyr::group_by(read_type) %>% 
  dplyr::mutate(x = dplyr::row_number())
```  

<!-- 
Melissa
# TODO THIS IS UGLY AND CAN'T SEE PATTERN
# TODO WHY IS NIFTIARRAY FUNCTION THE SAME MEMORY AS WRITENIFTIARRAY
-->

```{r}
ggplot(data=memory_profile, aes(x=x, y=log(memory_profile))) +
  geom_line()+
  geom_point() +
  facet_wrap(~read_type, scales = "free")
```

## Speed

Speed is very important. Obviously, we want code to be speed efficient to save overall computation time but as a user we also don't want to be distracted by time lags. 0.1 second is approximately the limit for a user to feel as though the system is reacting instantaneously. 1.0 second is about the limit for the user's flow or thought process to stay uninterrupted even though they notice a delay. 10 seconds is the limit to keep a users attention focused on the task at hand. The limit of 10 seconds a user will notice a delay and often lose their train of thought. Beyond 10 seconds and the user may lose track of even the task at hand. That is, they probably opened Twitter or Instagram and have completely lost track of what they were doing [Miller 1968; Card et al. 1991].

Therefore, speeds at 0.1 and 1 second limits are ideal to minimize lag and maximize user attention spans.

Card, S. K., Robertson, G. G., and Mackinlay, J. D. (1991). The information visualizer: An information workspace. Proc. ACM CHI'91 Conf. (New Orleans, LA, 28 April-2 May), 181-188.

Miller, R. B. (1968). Response time in man-computer conversational transactions. Proc. AFIPS Fall Joint Computer Conference Vol. 33, 267-277.

### One Patient Read

<!-- TODO: Explain the HDF5 and NIfTI NiftiArray functions. Maybe explain all the labels.-->

```{r}
speed = tibble::as_tibble(
  microbenchmark::microbenchmark(
    NiftiArray::NiftiArray(fileinfo$hdf5_destination[1]),
    NiftiArray::NiftiArray(fileinfo$nii_destination[1]),
    NiftiArray::writeNiftiArray(fileinfo$nii_destination[1]),
    RNifti::readNifti(fileinfo$nii_destination[1]),
    neurobase::readnii(fileinfo$nii_destination[1]),
    times = 5)
  ) %>% 
  dplyr::mutate(time = time/(1*10^9)) %>% # conver nanoseconds to seconds
  dplyr::rename(read_type = expr) %>% 
  dplyr::mutate(read_type = 
                  dplyr::case_when(
                    stringr::str_detect(read_type, 'RNifti') ~ "RNifti::readNifti",
                    stringr::str_detect(read_type, 'neurobase') ~ "neurobase::readnii",
                    stringr::str_detect(read_type, 'hdf5') ~ "NiftiArray::NiftiArray - HDF5 File",                    
                    stringr::str_detect(read_type, 'NiftiArray::NiftiArray\\(fileinfo\\$nii') ~ "NiftiArray::NiftiArray - NIfTI File",
                    stringr::str_detect(read_type, 'NiftiArray::writeNiftiArray') ~ "NiftiArray::writeNiftiArray"),
                read_type = as.factor(read_type),
                read_type = forcats::fct_reorder(read_type, time, .desc = TRUE))

# Table of memory
speed_summary = speed %>% 
  dplyr::group_by(read_type) %>% 
  dplyr::summarise(
    mean = mean(time, na.rm = TRUE),
    median = mean(time, na.rm = TRUE),
    sd = sd(time, na.rm = TRUE),
    min = min(time, na.rm = TRUE),
    max = max(time, na.rm = TRUE)
  )

knitr::kable(
  speed_summary,
  col.names = c('Read Function', 'Mean', 'Median', 'Std. Dev.', 'Min.', 'Max.'),
  format = 'html',
  digits = 2,
  caption = 'Memory mapped in bytes from a single patients image read.
  The image dimension is 182 by 218 by 182 with pixel dimension 1 mm by 1 mm by 1 mm.
  On disk the image is 4.9 MB.',
  booktabs = TRUE
) %>%
  kableExtra::kable_styling("striped", full_width = FALSE)
```

```{r}
ggplot(data = speed, aes(x = read_type, y = time)) +
  geom_boxplot() +   
  coord_flip() +
  ylab('Time (Seconds)') +
  xlab('Read Function') +
  labs(title = 'Speed Comparison for Reading a Single NIfTI Image',
       subtitle = 'Image Dimension: 182 by 218 by 182, Pixel Dimension: 1 mm by 1 mm by 1 mm') +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5), legend.position = 'none')
```

The `NiftiArray::NiftiArray - HDF5` and `RNifti::readNifti` both are around the 0.1 limit of seamless user flow. The remaining functions are around the 1 second limit where a user will notice a lag but not lose their train of thought.

<!--
TODO: Melissa
### Five Patient Read

repeat the single speed and memory stuff for 5 subjects.

TODO: Melissa
 
Repeat Speed but for writing not reading

# Suggested `NiftiArray` Workflow

Write pseudo code to do this

- convert all nifti images to hdf5 using writeNiftiArray
- load images using NiftiArrayList
- Do stuff across list or convert to matrix using acbind
- Calculate voxel level stuff maybe using DelayedMatrixStats or other stuff
- Convert back to an NiftiImage

TODO: 4D tutorial?

-->
