---
title: "NiftiArray: Fast Random Access of NIfTI Objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using NiftiArray}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE
)
```

This page is still under construction! Check back later for updates!

# Overview

`R` is not well suited for big datasets. NIfTI images, depending on the 
image dimension and voxel size, can be quite large when loaded and memory 
is a concern. As sample size increases it becomes difficult to perform simple
operations across subjects. For example, calculating the mean image across 
800 subjects is computationally intense since not all 800 subjects can be 
loaded into memory at once. Therefore, there is a need for alternative 
approaches. 

The Hierarchical Data Format version 5 (HDF5), is a file format that supports 
large, complex, heterogeneous data. This file format allows for fast random 
access of data on-disk. That is, HDF5 files can be directly manipulated without 
loading the object into memory and overcome a number of the issues related to 
working with big data in R. 

The `rhdf5` R package allows for direct manipulation of the HDF5 dataset on 
disk. This package can be utilized to convert, read, write, and manipulate 
the HDF5 data stored on disk. Other packages like `HDF5Array` provide 
memory-efficient on-disk access to the HDF5 dataset for standard R 
matrix/array manipulation. Simply put, as a user you feel as though 
you are working with a locally loaded large array (or matrix) and 
interact the same way you would if the object was loaded. For example, 
if you want to extract the value at array index 1,29,300 you can still 
simply call `data[1,29,.300]`. This large array `data` is never actually 
loaded into memory but rather is being accessed on disk in real time so 
memory is kept low and efficient.

Current methods were not designed for use explicitely with 3D array or 
images but support. So we extended and built something to seamlessly work 
with images....

DelayedArray tools...



The `NiftiArray` package allows for fast 
[random access](https://en.wikipedia.org/wiki/Random_access) 
of imaging data in NIfTI format and supports 
[`DelayedArray` operations](https://bioconductor.org/packages/release/bioc/html/DelayedArray.html). 
The pacakge establishes the `NiftiArray` class, a convenient and 
memory-efficient array-like container for on-disk representation of 
NIfTI image(s). The `NiftiArray` class is an extension of the 
[`HDF5Array` class](https://bioconductor.org/packages/release/bioc/html/HDF5Array.html) 
and converts NIfTI objects on disk to HDF5 files which allow for block 
processing and memory-efficient reprresentations in R.

This package allows users to save NIfTI objects as HDF5 files while 
conserving the array structure and NIfTI header. Once the data are 
stored as an HDF5 they can be represented as a `NIfTIArray` object. 
This object all

## Installation

You can install the development version of `NiftiArray` from
[GitHub](https://github.com/) with:

```{r}
# install.packages('remotes')
remotes::install_github("muschellij2/NiftiArray")
```

We are working to get a stable version on 
[Neuroconductor](www.neuroconductor.org).

## Packages

The packages you will need to load for use with this tutorial are below:

```{r}
library(NiftiArray)
```


